Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-06 paper/paper.tex
compressed cplx-real-calculus review in appendix

Complex networks are, in general, non-holomorphic, i.e. not $\cplx$-differentiable, which is
exacerbated by the loss being real-valued. Indeed, as implied by Cauchy-Riemann conditions, a
$\cplx$-differentiable real-valued function is necessarily trivial. This issue is dealt with
by employing Wirtinger, or $\cplx\real$ calculus, which generalizes holomorphic calculus
to non-holomorphic functions of $\cplx$-argument, \cite{adali_complex-valued_2011}. It was
considered as a basis for $\cplx$ version of backpropagation in \cite{benvenuto_complex_1992}
and \cite{trabelsi_deep_2017} (see sec.~\ref{sub:wirtinger_calculus} for details). The key
benefit of $\cplx\real$ calculus is straightforward symbolic differentiation, and the fact,
that this allows retrofiting $\cplx$-valued networks into existing $\real$ deep learning
auto-differentiation libraries.

@@

% TLDR: we use Wirtinger calculus!
Basically $\cplx\real$ calculus regard $f(z)$ as a function of two independent variables $z$ 
and its conjugate $\conj{z}$ and symbolically defines derivative operators $
  \tfrac{\partial}{\partial z}
$ and $
  \tfrac{\partial}{\partial \conj{z}}
$, that satisfy linearity, product and chain rules, and respect complex conjugation. The important
property is that, if we identify $f(z) = f(u + jv) = F(u, v)$, then the $\cplx\real$ calculus
rules differential of $f$ as a $\cplx$ function coincides its differential as an $\real^2$
function. This makes it possible to use and apply the rules of $\real$-calculus to the networks
represented in paired $\real$ fashion and use existing $\real$-valued backpropagation in autograd
software. The key analytic advantage is that computing derivatives is more streamlined, less
messy, and derivative expressions involving complex variables turn out to be simpler that their
real-imaginary analogues, \cite{adali_complex-valued_2011}.

% essential intro into Wirtinger calculus (CR)
% https://math.stackexchange.com/a/444493
Wirtinger calculus relies on the natural identification of $\cplx$ with $\real \times \real$.
Let's consider $
  f\colon \cplx \to \cplx
$ and treat it as an equivalent function $\real^2 \to \cplx$ via $
  f(z) = f(u + jv) = F(u, v)
$. Wirtinger calculus defines two formal derivative operators
$
  \tfrac{\partial}{\partial z}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      - j \tfrac{\partial}{\partial v}
    \bigr)
$ and $
  \tfrac{\partial}{\partial \conj{z}}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      + j \tfrac{\partial}{\partial v}
    \bigr)
$ and the differential of $f$ at $z \in \cplx$ as
$$
df(z) \big\vert_{z=u + jv}
  = \frac{\partial f}{\partial z} dz
    + \frac{\partial f}{\partial \conj{z}} d\conj{z}
   % = \frac12\bigl(
   %    \frac{\partial}{\partial u}
   %    - j \frac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % + \frac12\bigl(
   %    \frac{\partial}{\partial u}
   %    + j \frac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % = \frac12 \bigl(
   %    \frac{\partial F}{\partial u} du - j \frac{\partial F}{\partial v} du
   %    + j \frac{\partial F}{\partial u} dv + \frac{\partial F}{\partial v} dv
   % \bigr)
   % + \frac12 \bigl(
   %    \frac{\partial F}{\partial u} du + j \frac{\partial F}{\partial v} du
   %    - j \frac{\partial F}{\partial u} dv + \frac{\partial F}{\partial v} dv
   % \bigr)
   = \frac{\partial F}{\partial u} du
     + \frac{\partial F}{\partial v} dv
   = dF(u, v)
  \,, $$
where $dz = du + j dv$ and $d\conj{z} = du - j dv$. This implies that the complex argument
and its conjugate are treated as independent variables. In this notation Cauchy-Riemann
conditions are expressed as $
  \tfrac{\partial f}{\partial \conj{z}} = 0
$, or $
  -j \tfrac{\partial F}{\partial v} = \tfrac{\partial F}{\partial v}
$. These conditions impose a rigid structure on real and imaginary parts of $F(u, v)$, namely $
  \tfrac{\partial F_{\Re }}{\partial u} = \tfrac{\partial F_{\Im }}{\partial v}
$ and $
  \tfrac{\partial F_{\Re }}{\partial v} = - \tfrac{\partial F_{\Im }}{\partial u}
$. Thus Wirtinger calculus subsumes the usual $\cplx$-calculus on holomorphic functions,
when $
  \tfrac{\partial f}{\partial \conj{z}} = 0
$, i.e. $
  f(z)
    % = g(z, \conj{z})
    = F(\tfrac12 (z + \conj{z}), \tfrac1{j 2} (z - \conj{z}))
$ does not depend on $\conj{z}$. The usual rules of calculus, like chain and product rules,
follow directly from the definition of the operators, e.g.
$$
  \frac{\partial (f\circ g)}{\partial z}
    = \frac{\partial f(g(z))}{\partial g} \frac{\partial g(z)}{\partial z}
    + \frac{\partial f(g(z))}{\partial \conj{g}} \frac{\partial \conj{g(z)}}{\partial z}
    % = \nabla G \nabla F
  \,. $$
% Show that $dh(z) = df(g(z)) dg(z)$.
% $h(z) = f(g(z)) = F(G_r(u, v), G_i(u, v)) = H(u, v)$
% $$
% dH
%   = \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial u} du
%   + \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial v} dv
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial u} du
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial v} dv
%   = \tfrac{\partial F}{\partial r} d\Re g(z)
%   + \tfrac{\partial F}{\partial i} d\Im g(z)
%   = df(\Re g(z) + j\Im g(z)) d g(z)
%   = df(g(z)) d g(z)
%   \,. $$

Since in machine learning the target loss is typically an empirical real-valued risk, which
is minimized, and we let's derive the expression for the gradient and make sure that it is
compatible with real-valued calculus upon which current auto-differentiation tools are built.
For a real-valued $f\colon \cplx \to \real$ we have $\conj{f} = f$, whence $
  \tfrac{\partial f}{\partial \conj{z}}
    = \tfrac{\partial \conj{f}}{\partial \conj{z}}
    = \conj{\tfrac{\partial f}{\partial z}}
$. Hence
$$
df
  = \tfrac{\partial f}{\partial z} dz
    + \tfrac{\partial f}{\partial \conj{z}} d\conj{z}
  % = \conj{\tfrac{\partial \conj{f}}{\partial \conj{z}}} dz
  % + \conj{\conj{\tfrac{\partial f}{\partial \conj{z}}} dz}
  = 2 \Re \Bigl(
    \conj{\tfrac{\partial f}{\partial \conj{z}}} dz
  \Bigr)
  = 2 \Re \bigl(
    \tfrac{\partial f}{\partial z} dz
  \bigr)
  \,. $$
Hence the gradient of $f$ at $z$ is given by $
  \tfrac{\partial f}{\partial \conj{z}}
    = \tfrac{\partial F}{\partial u}
      + j \tfrac{\partial F}{\partial v}
$, meaning that we can use gradients with respect ral and imaginary parts independently. Thus
using Wirtinger calculus makes it possible to reuse $\real$-backpropagation for $\cplx$-networks.

>>>>

Complex networks are, in general, non-holomorphic, i.e. not $\cplx$-differentiable, which is
exacerbated by the loss being real-valued. Indeed a $\cplx$-differentiable real-valued function is
necessarily trivial by Cauchy-Riemann conditions. This issue is dealt with by employing Wirtinger,
or $\cplx\real$ calculus, which generalizes holomorphic calculus to non-holomorphic functions
of $\cplx$-argument, \cite{adali_complex-valued_2011}. The rules it defines satisfy intuitive
linearity, product and chain rules, and respect complex conjugation, but most importantly the
$\cplx\real$-differential of a function on $\cplx$ coincides with its classical differential as
a function on $\real^2$. Essentially, it allows straightforward retrofiting of $\cplx$-valued
networks into existing $\real$ deep learning auto-differentiation frameworks. It was considered
as a basis for $\cplx$ version of backpropagation in \cite{benvenuto_complex_1992} and
\cite{trabelsi_deep_2017}, (see sec.~\ref{sub:wirtinger_calculus} for details).

@@

% TLDR: we use Wirtinger calculus!

% essential intro into Wirtinger calculus (CR)
% https://math.stackexchange.com/a/444493
Wirtinger ($\cplx\real$) calculus relies on the natural identification of $\cplx$ with $
  \real \times \real
$, and regards $
  f\colon \cplx \to \cplx
$ as an equivalent in algebraic sense function $F\colon \real^2 \to \cplx$ defined $
  f(z) = f(u + jv) = F(u, v)
$. Under this identification, we define two formal derivative operators
$
  \tfrac{\partial}{\partial z}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      - j \tfrac{\partial}{\partial v}
    \bigr)
$ and $
  \tfrac{\partial}{\partial \conj{z}}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      + j \tfrac{\partial}{\partial v}
    \bigr)
$ and the differential of $f$ at $z \in \cplx$ as
$$
df(z) \big\vert_{z=u + jv}
  = \frac{\partial f}{\partial z} dz
    + \frac{\partial f}{\partial \conj{z}} d\conj{z}
   % = \frac12\bigl(
   %    \frac{\partial}{\partial u}
   %    - j \frac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % + \frac12\bigl(
   %    \frac{\partial}{\partial u}
   %    + j \frac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % = \frac12 \bigl(
   %    \frac{\partial F}{\partial u} du - j \frac{\partial F}{\partial v} du
   %    + j \frac{\partial F}{\partial u} dv + \frac{\partial F}{\partial v} dv
   % \bigr)
   % + \frac12 \bigl(
   %    \frac{\partial F}{\partial u} du + j \frac{\partial F}{\partial v} du
   %    - j \frac{\partial F}{\partial u} dv + \frac{\partial F}{\partial v} dv
   % \bigr)
   = \frac{\partial F}{\partial u} du
     + \frac{\partial F}{\partial v} dv
   = dF(u, v)
  \,, $$
where $dz = du + j dv$ and $d\conj{z} = du - j dv$. This implies that the complex argument
and its conjugate are treated as independent variables. Cauchy-Riemann conditions $
  -j \tfrac{\partial F}{\partial v} = \tfrac{\partial F}{\partial u}
$ impose a rigid structure on real and imaginary parts of $F$ and can be expressed as $
  \tfrac{\partial f}{\partial \conj{z}} = 0
$ in this notation. Thus $\cplx\real$ calculus subsumes the usual $\cplx$-calculus on
holomorphic functions, when $f(z)$, regarded as $f(z, \conj{z})$, is constant with respect
to $\conj{z}$. The usual rules of calculus, like chain and product rules, follow directly
from the definition of the operators, e.g.
$$
  \frac{\partial (f\circ g)}{\partial z}
    = \frac{\partial f(g(z))}{\partial g} \frac{\partial g(z)}{\partial z}
    + \frac{\partial f(g(z))}{\partial \conj{g}} \frac{\partial \conj{g(z)}}{\partial z}
    % = \nabla G \nabla F
  \,. $$
% Show that $dh(z) = df(g(z)) dg(z)$.
% $h(z) = f(g(z)) = F(G_r(u, v), G_i(u, v)) = H(u, v)$
% $$
% dH
%   = \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial u} du
%   + \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial v} dv
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial u} du
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial v} dv
%   = \tfrac{\partial F}{\partial r} d\Re g(z)
%   + \tfrac{\partial F}{\partial i} d\Im g(z)
%   = df(\Re g(z) + j\Im g(z)) d g(z)
%   = df(g(z)) d g(z)
%   \,. $$

In machine learning tasks the target objective is typically empirical risk, which has to
be real-valued to be minimized. Nevertheless, the expression of the $\cplx\real$ gradient
is compatible with what is expected, when $f$ is treated like a $\real^2$ function. For a
real-valued $f\colon \cplx \to \real$ we have $\conj{f} = f$, which implies $
  \tfrac{\partial f}{\partial \conj{z}}
    = \tfrac{\partial \conj{f}}{\partial \conj{z}}
    = \conj{\tfrac{\partial f}{\partial z}}
$, whence
$$
df
  = \tfrac{\partial f}{\partial z} dz
    + \tfrac{\partial f}{\partial \conj{z}} d\conj{z}
  % = \conj{\tfrac{\partial \conj{f}}{\partial \conj{z}}} dz
  % + \conj{\conj{\tfrac{\partial f}{\partial \conj{z}}} dz}
  = 2 \Re \Bigl(
    \conj{\tfrac{\partial f}{\partial \conj{z}}} dz
  \Bigr)
  = 2 \Re \bigl(
    \tfrac{\partial f}{\partial z} dz
  \bigr)
  \,. $$
Thus, the gradient of $f$ at $z$ is given by $
  \nabla_z f(z)
    = \tfrac{\partial F}{\partial u}
      + j \tfrac{\partial F}{\partial v}
$. Therefore, Wirtinger calculus makes it possible to reuse $\real$-backpropagation.
